Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-21 paper/supplementary.tex
maciej's edit
change colors, setup subfigure caption, add annotation to tsne
Updated paper.

Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label after each prediction.
Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
\begin{align}
    \mathrm{fb}[t]=\left\{\begin{array}{cl}
    0 & : t \in\{1, \ldots, B\} \\
    B & : t \in\{B+1, \ldots, 2 B\} \\
    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
    & \vdots
    \end{array}\right.
\end{align}

@@

\subsection{Machine Learning Design Pipeline}
In this section, we report in details about our machine learning design pipeline.
The raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.

@@

\textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.
We first normalise the raw TIR to values between 0 and 1. 
We applied the Gaussian Process Regression with noise parameter $\alpha = 1e-10$. 
We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
The design size is 60 with UCB parameter $\beta = 1$.\\
\textbf{Bandit-1 to Bandit-3: }
the t + 1 round design is based on the $t^th$ ($t \geq 1$) round result, where each sequence has 6 replicates with TIR labels.
We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively.
After normalisation, each replicate has zero mean and unit variance. 
For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
We model the covariance matrix using the weighted degree kernel with shift. 
We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation. 
We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
We recommend 90 sequences from the design space.

@@

In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS. 
This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.

@@

@@

>>>>

Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label after each prediction.
Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2014parallelizing}.
With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
\begin{align}
    \mathrm{fb}[t]=\left\{\begin{array}{cl}
    0 & : t \in\{1, \ldots, B\} \\
    B & : t \in\{B+1, \ldots, 2 B\} \\
    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
    & \vdots
    \end{array}\right.
\end{align}

@@

\subsection{Machine Learning Design Pipeline}
\label{sec: ML design pipeline}
In this section, we report in details about our machine learning design pipeline.
The raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.

@@

\textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.
We first normalise the raw TIR to values between 0 and 1. 
We applied the Gaussian Process Regression with noise parameter $\alpha = 1e-10$. 
We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
The design size is 60 with UCB parameter $\beta = 1$.\\
\textbf{Bandit-1 to Bandit-3: }
the t + 1 round design is based on the $t^th$ ($t \geq 1$) round result, where each sequence has 6 replicates with TIR labels.
We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively.
After normalisation, each replicate has zero mean and unit variance. 
For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
We model the covariance matrix using the weighted degree kernel with shift. 
We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation. 
We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2014parallelizing}. 
The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
We recommend 90 sequences from the design space.

@@

In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS. 
This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.

@@

\subsection{Other Evaluation}

\textbf{How much more efficient we are by using GP-UCB?}

random = {set of TIR that are in UNI and PPM}
bandit = {set of TIR that are in bandit 0,1,2,3}

The ratio of the max random to the max bandit: 0.67/1.34 = 0.5
The ratio of the 0.9 quantile random to the 0.9 quantile bandit: 0.49/0.82 = 0.598

we can also calculate an average regret of TIRs which are higher than SD? The regret for one sequences is calculated based on the difference between the benchmark (1) and TIR.

reciprocal would make a better story. I.e. bandit is 2x better than random.

\textbf{zoom-in tSNE?}

1) supplementary website.
2) eLife is now offering submissions of interactive papers with Jupyter notebooks. More of a trivia than useful information in our case.
https://elifesciences.org/labs/ad58f08d/introducing-elife-s-first-computationally-reproducible-article

@@

\begin{figure}[!ht]
    \centering
     \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF.pdf}
    \caption{Exploration v.s. Exploitation.}
\end{figure}

\begin{figure}[!ht]
    \centering
     \includegraphics[scale=0.4]{paper/plots/Supplementary/SDhist.pdf}
    \caption{Histogram of standard deviations.}
\end{figure}

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \caption{}
         \includegraphics[scale=0.4]{plots/Supplementary/SD.pdf}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \caption{}
         \includegraphics[scale=0.6]{plots/Supplementary/SDvsTIR.pdf}
     \end{subfigure}
     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation). B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR. Standard deviation and average TIR is given in terms of raw number.}
     \label{fig:variation of biological replicates}
\end{figure}
